# Metadata QC

Reconcile the IMAGINE metadata with the Surette sample data sheet.

```{r}
library(tidyverse)
```

## IMAGINE metadata

Reading in the IMAGINE metadata sheet I received from Aida, 2025-09-29.

```{r}
list.files('input/')
img_meta = read.csv(file = 'input/active_IMAGINE_metadata_wide.csv', 
                    strip.white = TRUE)
head(img_meta)

meta_long = (img_meta
             %>% pivot_longer(Baseline:Y4, names_to = 'Timepoint',
                              values_to = 'Sample.ID'))

head(meta_long)

meta_long = filter(meta_long,!is.na(Sample.ID))
dim(meta_long)
```

This file has 13,516 samples in it. I'm going to clean it up a bit:

```{r}
head(meta_long)
summary(meta_long)

meta_long = (meta_long
             %>% mutate(IBD.Classification = case_when(Disease.Type %in% c('IBS',
                                                               'Healthy') ~ 
                                                       Disease.Type,
                                                   TRUE ~ IBD.Classification),
                    IBD.Classification = factor(IBD.Classification,
                                                levels = c('CD', 'UC',
                                                           'IBD unclassified',
                                                          'IBS', 'Healthy','')),
                    Timepoint = factor(Timepoint, levels = c('Baseline',
                                                             'Y1','Y2','Y3',
                                                             'Y4')),
                    Disease.Type = factor(Disease.Type, levels = c('Healthy',
                                                                   'IBS','IBD',
                                                                   '')),
                    across(c(Site.Number, Participant.ID, Sample.ID), factor)))
```

Laura confirmed by email that sample 203664 was destroyed, so it should be
removed from the listing to prevent confusion during reconciliation:

```{r}
meta_long = filter(meta_long, Sample.ID != 203664)
dim(meta_long)
```

Now we have 13,515 samples.

```{r}
summary(meta_long)
```

It doesn't look like there are any duplicate sample IDs. Let's make sure:

```{r}
meta_dup = (meta_long
            %>% count(Sample.ID)
            %>% filter(n > 1)
            %>% left_join(meta_long, by = 'Sample.ID'))
dim(meta_dup)
meta_dup
```

No need to write anything to a file
```{r}
# write.csv(meta_dup,'intermed/IMAGINE_duplicate_IDs_YYYY-MM-DD.csv')
```

## Surette Lab Sample Info Sheet

Now I am loading the most recent version of Laura's sample info sheet

```{r}
samp_inf = read.csv(file = 'input/active_Surette_sample_info_sheet.csv',
                    strip.white = TRUE, row.names = 1)
dim(samp_inf)
```

This is 13,526 samples. That is 11 more samples than in the IMAGINE metadata
sheet. I will clean this up and remove any columns I don't need:

```{r}
summary(samp_inf)
head(samp_inf)

samp_inf = (samp_inf
            %>% select(Sample.ID, Study.ID, Date.Received.YYYY.MM.DD, 
                       Illumina.Plate.Number, Notes)
            %>% rename(Date.Received = Date.Received.YYYY.MM.DD)
            %>% mutate(across(c(Sample.ID, Study.ID, Illumina.Plate.Number),
                              factor),
                       Date.Received = dmy_hm(Date.Received)))
```

Tania confirmed by email that sample 101246 should be removed from analysis,
since it cannot be associated with a patient/timepoint. We are also removing
sample 200510 because it is a duplicate of 100785.

```{r}
samp_inf = filter(samp_inf, !(Sample.ID %in% c(101246, 200510)))
dim(samp_inf)
```

There are now 13,524 samples, 9 more than IMAGINE.

```{r}
summary(samp_inf)
```

I don't see any duplicates, but let's double check:

```{r}
samp_dup = (samp_inf
            %>% count(Sample.ID)
            %>% filter(n > 1)
            %>% left_join(samp_inf, by = 'Sample.ID'))
dim(samp_dup)
samp_dup
```

Excellent.

This step is not currently necessary, because there are no duplicates:

```{r}
# any(meta_dup$Sample.ID %in% samp_dup$Sample.ID)
```

No duplicates to write:

```{r}
# write.csv(samp_dup, 
#           file = 'intermed/Surette_sample_info_duplicate_SampleIDs.csv')
```

Check if there are any duplicated Study IDs:

```{r}
samp_inf %>% count(Study.ID) %>% filter(n > 1)
```

Nope, none. Excellent!

This step compares duplicates between sheets, but is not necessary because there
are no duplicates.

```{r}
# any(meta_dup$Sample.ID %in% samp_inf$Sample.ID)
# all(meta_dup$Sample.ID %in% samp_inf$Sample.ID)
```

## Compare the two sheets

Are all the metadata Sample IDs in the sample info sheet?

```{r}
all(meta_long$Sample.ID %in% samp_inf$Sample.ID)
meta_only = filter(meta_long, !(Sample.ID %in% samp_inf$Sample.ID))
dim(meta_only)
```

There is one sample ID in the IMAGINE metadata sheet that is absent from the
Surette lab sheet:

```{r}
data.frame(meta_only)
```

This is sample 100446. I wrote to the team to ask about it on October 17, 2025,
and followed up on December 5, 2025. I am wondering if 100446 is a typo for
101246, which would mean we could re-include that sample.

```{r}
write.csv(meta_only, 
          file = 'intermed/IMAGINE_samps_missing_from_Surette_lab_sheet.csv')
```

Okay, what about the other way around? What is in the sample info sheet but
missing from the IMAGINE metadata?

```{r}
all(samp_inf$Sample.ID %in% meta_long$Sample.ID)
samp_only = filter(samp_inf, !(Sample.ID %in% meta_long$Sample.ID))
dim(samp_only)
```

As of December 5, 2025, there are 10 samples in the Surette sample data sheet
that are missing from the IMAGINE sheet.

```{r}
samp_only
samp_only %>% filter(grepl('D', Study.ID) |
                         grepl('discard', Notes, ignore.case = TRUE) |
                         grepl('ineligible', Notes, ignore.case = TRUE) |
                         grepl('dispose', Notes, ignore.case = TRUE))
```

9 of these are discards, leaving 1:

```{r}
samp_only = (samp_only 
              %>% filter(!grepl('D', Study.ID), 
                         !grepl('discard', Notes, ignore.case = TRUE),
                         !grepl('ineligible', Notes, ignore.case = TRUE)))
samp_only
```

We are waiting to hear back about this one from Paul.

```{r}
write.csv(samp_only, 
          file = 'intermed/SuretteLab_samps_missing_from_IMAGINE_sheet.csv')
```

# Sequencing QC

This is where I check for re-sequenced samples and merge them.

```{r}
asvtab = readRDS('input/active_mergetab_nochim.rds')
dim(asvtab)
asvtab = t(asvtab)
dim(asvtab)
```

There are 13,125 samples sequenced.

```{r}
seq_ids = colnames(asvtab)
head(seq_ids)
negs = seq_ids[!startsWith(seq_ids, 'IMG')]
length(negs)
negs
```

There are 70 negative controls. I will partition them off for now

```{r}
neg_ctl = asvtab[,negs]
seq_rem = seq_ids[!(seq_ids %in% negs)]
asvtab_no_neg = asvtab[,seq_rem]
dim(asvtab_no_neg)
```

This leaves 13,055 samples. Now let's look for any IMG ids that are missing from
the sample data sheet

```{r}
seq_only = seq_rem[!(seq_rem %in% samp_inf$Study.ID)]
length(seq_only)
sort(seq_only)
```

There are 49 samples that start with IMG but are missing from the sample info
sheet. Most of these end in A or B, which means they are re-sequenced samples,
but a couple don't. Let's look at those:

```{r}
asv_missing = seq_only[!grepl('[AB]$',seq_only)]
length(asv_missing)
asv_missing

samp_inf %>% filter(grepl('D', Study.ID) |
                         grepl('discard', Notes, ignore.case = TRUE) |
                         grepl('ineligible', Notes, ignore.case = TRUE) |
                         grepl('dispose', Notes, ignore.case = TRUE))
```

There are 7 sequenced samples missing from the sample info sheet. All of these
are listed in the sample info sheet as discard/DNP, so I will remove them from
the final table. This includes IMG8674, which is not currently listed in the
sample info sheet as a discard, but which Tania and Laura have confirmed should
be discarded, as well as IMG3919, which is sample 200510 and is a duplicate.

Now I'm going to check that all the duplicated samples are present in their
original forms in the asv table

```{r}
dups = seq_only[grepl('[AB]$', seq_only)]
length(dups)
dups

dup_origs = substr(dups, 1, nchar(dups) -1)
dup_origs
n_distinct(dups)
n_distinct(dup_origs)

all(dup_origs %in% seq_rem)
```

Nope! Some are missing!

```{r}
sum(dup_origs %in% seq_rem)
```

Only 12 are in there.

```{r}
dup_origs[dup_origs %in% colnames(asvtab)]
dup_origs[!(dup_origs %in% colnames(asvtab))]
```

This is confusing. Let's see what's up with these guys in samp_info:

```{r}
sum(dup_origs %in% samp_inf$Study.ID)
```

All of them are in sample info, which is promising. Is it an A vs B thing?

```{r}
all(dup_origs[endsWith(dups, 'A')] == 
        dup_origs[!(dup_origs %in% colnames(asvtab))])
```

Okay, so the ones that end in A are lacking originals in the ASV table, and the
ones that end in B are actual duplicates that need to be merged. Cool.

```{r}
reseq = dups[endsWith(dups, 'A')]
reseq_orig = dup_origs[endsWith(dups, 'A')]

dup_origs = dup_origs[endsWith(dups, 'B')]
dups = dups[endsWith(dups, 'B')]
```

Remove the actual duplicates to their own table for now

```{r}
duptab = asvtab_no_neg[,c(dups, dup_origs)]
dim(duptab)
seq_rem = seq_rem[!(seq_rem %in% c(dups, dup_origs))]
length(seq_rem)
asvtab_no_dup = asvtab_no_neg[, seq_rem]
```

Remove the A from the column names of the non-duplicated samples so they match
the metadata

```{r}
colnames(asvtab_no_dup) = str_replace(colnames(asvtab_no_dup), 'A','')
seq_rem = colnames(asvtab_no_dup)
```

Merge the sequences of the duplicated samples. Make sure the IDs actually match.

```{r}
data.frame(dups, dup_origs)
```

They match, so we'll iterate through it:

```{r}
dedupped = matrix(nrow = nrow(duptab), ncol = length(dups))
for (i in 1:length(dups)){
    d = dups[i]
    o = dup_origs[i]
    dedupped[,i] = rowSums(duptab[,c(d,o)])
}
head(dedupped)
colnames(dedupped) = dup_origs
rownames(dedupped) = rownames(duptab)
```

Bring all the samples back together:

```{r}
asvtab_fixed = cbind(neg_ctl, asvtab_no_dup, dedupped)
dim(asvtab_fixed)
```

Let's double check that adding up the samples worked:

```{r}
asv_tots = rowSums(asvtab)
asv_tot_dedup = rowSums(asvtab_fixed)
all(asv_tots == asv_tot_dedup)

dim(asvtab)
dim(asvtab_fixed)
length(dups)
```

The rowsums are the same and there are 12 samples fewer in the fixed table than
the original table, which is the right number of duplicates. 

How many of the samples are missing from the sample info sheet?

```{r}
samp_ids = colnames(asvtab_fixed)
head(samp_ids)
tail(samp_ids)
head(samp_inf)
missing = samp_ids[!(samp_ids %in% c(samp_inf$Study.ID))]
missing = missing[!(missing %in% negs)]
missing
```

Just the 7 that should be discarded. I'll discard them now.

```{r}
dim(asvtab_fixed)
asvtab_fixed = asvtab_fixed[,colnames(asvtab_fixed)[!(colnames(asvtab_fixed) %in% 
                                                          missing)]]
dim(asvtab_fixed)
```

Removed 7 samples.

Removing a sample may bring some ASVs down to 0 everywhere. Let's check and
remove them if necessary:

```{r}
dim(asvtab_fixed)
sum(rowSums(asvtab_fixed) == 0)
asvtab_fixed = asvtab_fixed[rowSums(asvtab_fixed) > 0,]
dim(asvtab_fixed)
```

There were 43 ASVs with 0s everywhere. They have been removed, leaving us with 
103,096 ASVs in 13,106 samples.

# Write Fixed Tables

Create a maptab that is merged from the IMAGINE and Surette metadat sheets and
an asvtab where all the duplicates are merged, the sample IDs match the maptab,
and the discards are removed.

```{r}
maptab = full_join(samp_inf, meta_long, by = 'Sample.ID')
dim(maptab)
```

The negative controls aren't in the maptab, which means they will be removed
from everything downstream of this once we start using phyloseq. I'll put them
in.

```{r}
head(maptab)
plate = unlist(strsplit(negs, 'n'))[(1:length(negs)) * 2 -1]
plate
neg_map = data.frame(Sample.ID = NA,
                     Study.ID = negs,
                     Date.Received = NA,
                     Illumina.Plate.Number = plate,
                     Notes = 'negative control',
                     Site.Number = NA,
                     Participant.ID = NA,
                     Disease.Type = NA,
                     IBD.Classification = NA,
                     Timepoint = NA)
maptab = rbind(maptab, neg_map)
dim(maptab)
```

Including negative controls, that leaves us with 13,595 samples in the maptab,
of which 13,106 are present in the ASV table I currently have.

```{r}
write_rds(maptab, file = 'data/merged_maptab.rds')
write_rds(asvtab_fixed, file = 'data/merged_seqtab.rds')
```

Start tracking sample/taxon counts

```{r}
dim(maptab)
dim(asvtab_fixed)

stat_df = data.frame(Step = '00_manual_qc_steps.Rmd',
                     Samples = c(nrow(maptab), ncol(asvtab_fixed)),
                     Taxa = c(NA, ncol(asvtab_fixed)),
                     File = c('data/merged_maptab.rds', 
                              'data/merged_seqtab.rds'))

if (!dir.exists('stats')){
    dir.create('stats')
}

write.csv(stat_df, file = 'stats/track_counts.csv', row.names = FALSE)
```

