# Metadata QC

Reconcile the IMAGINE metadata with the Surette sample data sheet.

```{r}
library(tidyverse)
```

## IMAGINE metadata

Reading in the IMAGINE metadata sheet I received from Aida, 2025-09-29.

```{r}
list.files('input/')
img_meta = read.csv(file = 'input/active_IMAGINE_metadata_wide.csv', 
                    strip.white = TRUE)
head(img_meta)

meta_long = (img_meta
             %>% pivot_longer(Baseline:Y4, names_to = 'Timepoint',
                              values_to = 'Sample.ID'))

head(meta_long)

meta_long = filter(meta_long,!is.na(Sample.ID))
dim(meta_long)
```

This file has 13,516 samples in it. I'm going to clean it up a bit:

```{r}
head(meta_long)
summary(meta_long)

meta_long = (meta_long
             %>% mutate(IBD.Classification = case_when(Disease.Type %in% c('IBS',
                                                               'Healthy') ~ 
                                                       Disease.Type,
                                                   TRUE ~ IBD.Classification),
                    IBD.Classification = factor(IBD.Classification,
                                                levels = c('CD', 'UC',
                                                           'IBD unclassified',
                                                          'IBS', 'Healthy','')),
                    Timepoint = factor(Timepoint, levels = c('Baseline',
                                                             'Y1','Y2','Y3',
                                                             'Y4')),
                    Disease.Type = factor(Disease.Type, levels = c('Healthy',
                                                                   'IBS','IBD',
                                                                   '')),
                    across(c(Site.Number, Participant.ID, Sample.ID), factor)))
```

Laura confirmed by email that sample 203664 was destroyed, so it should be
removed from the listing to prevent confusion during reconciliation:

```{r}
meta_long = filter(meta_long, Sample.ID != 203664)
dim(meta_long)
```

Now we have 13,515 samples in the IMAGINE metadata file.

```{r}
summary(meta_long)
```

It doesn't look like there are any duplicate sample IDs. Let's make sure:

```{r}
meta_dup = (meta_long
            %>% count(Sample.ID)
            %>% filter(n > 1)
            %>% left_join(meta_long, by = 'Sample.ID'))
dim(meta_dup)
meta_dup
```

No duplicates; no need to write anything to a file
```{r}
# write.csv(meta_dup,'intermed/IMAGINE_duplicate_IDs_YYYY-MM-DD.csv')
```

## Surette Lab Sample Info Sheet

Now I am loading the most recent version of Laura's sample info sheet

```{r}
samp_inf = read.csv(file = 'input/active_Surette_sample_info_sheet.csv',
                    strip.white = TRUE, row.names = 1)
dim(samp_inf)
```

This is 13,526 samples. That is 11 more samples than in the IMAGINE metadata
sheet. I will clean this up and remove any columns I don't need:

```{r}
summary(samp_inf)
head(samp_inf)

samp_inf = (samp_inf
            %>% select(Sample.ID, Study.ID, Date.Received.YYYY.MM.DD, 
                       Illumina.Plate.Number, Notes)
            %>% rename(Date.Received = Date.Received.YYYY.MM.DD)
            %>% mutate(across(c(Sample.ID, Study.ID, Illumina.Plate.Number),
                              factor),
                       Date.Received = dmy_hm(Date.Received)))
```

Tania confirmed by email that sample 101246 should be removed from analysis,
since it cannot be associated with a patient/timepoint. We are also removing
sample 200510 because it is a duplicate of 100785.

```{r}
samp_inf_rm = c()
samp_inf_rm = c(samp_inf_rm,
               (samp_inf %>% filter(Sample.ID %in% c(101246,200510)) 
                %>% pull(Study.ID)
                %>% as.character))
samp_inf = filter(samp_inf, !(Sample.ID %in% c(101246, 200510)))
dim(samp_inf)
```

There are now 13,524 samples in the Surette sample info sheet, 9 more than
IMAGINE.

```{r}
summary(samp_inf)
```

I don't see any duplicates, but let's double check:

```{r}
samp_dup = (samp_inf
            %>% count(Sample.ID)
            %>% filter(n > 1)
            %>% left_join(samp_inf, by = 'Sample.ID'))
dim(samp_dup)
samp_dup
```

Excellent.

This step is not currently necessary, because there are no duplicates:

```{r}
# any(meta_dup$Sample.ID %in% samp_dup$Sample.ID)
```

No duplicates to write:

```{r}
# write.csv(samp_dup, 
#           file = 'intermed/Surette_sample_info_duplicate_SampleIDs.csv')
```

Check if there are any duplicated Study IDs:

```{r}
samp_inf %>% count(Study.ID) %>% filter(n > 1)
```

Nope, none. Excellent!

This step compares duplicates between sheets, but is not necessary because there
are no duplicates.

```{r}
# any(meta_dup$Sample.ID %in% samp_inf$Sample.ID)
# all(meta_dup$Sample.ID %in% samp_inf$Sample.ID)
```

## Compare the two sheets

Are all the metadata Sample IDs in the sample info sheet?

```{r}
all(meta_long$Sample.ID %in% samp_inf$Sample.ID)
meta_only = filter(meta_long, !(Sample.ID %in% samp_inf$Sample.ID))
dim(meta_only)
```

There is one sample in the IMAGINE sheet that is absent from the Surette lab
sheet

```{r}
data.frame(meta_only)
```

We confirmed by email in December of 2025 that '100446' could not be
identified and should be omitted from analysis.

```{r}
meta_long = filter(meta_long, Sample.ID != '100446')
dim(meta_long)
```

This leaves us with 13,514 samples in the IMAGINE metadata.

Okay, what about the other way around? What is in the sample info sheet but
missing from the IMAGINE metadata?

```{r}
all(samp_inf$Sample.ID %in% meta_long$Sample.ID)
samp_only = filter(samp_inf, !(Sample.ID %in% meta_long$Sample.ID))
dim(samp_only)
```

As of April 24, 2026, there are 10 samples in the Surette sample data sheet
that are missing from the IMAGINE sheet.

```{r}
samp_only
samp_only %>% filter(grepl('D', Study.ID) |
                         grepl('discard', Notes, ignore.case = TRUE) |
                         grepl('ineligible', Notes, ignore.case = TRUE) |
                         grepl('dispose', Notes, ignore.case = TRUE))
```

9 of these are discards, leaving 1:

```{r}
samp_only = (samp_only 
              %>% filter(!grepl('D', Study.ID), 
                         !grepl('discard', Notes, ignore.case = TRUE),
                         !grepl('ineligible', Notes, ignore.case = TRUE)))
samp_only
```

Paul Moayyedi confirmed by email that 202410 should be excluded from analysis

```{r}
samp_inf_rm = c(samp_inf_rm,
               (samp_inf %>% filter(Sample.ID == 202410) 
                %>% pull(Study.ID)
                %>% as.character))
samp_inf = filter(samp_inf, Sample.ID != '202410')
```

This is everything. There are no further discrepancies between the two datasets!

# Sequencing QC

This is where I check for re-sequenced samples and merge them.

```{r}
asvtab = readRDS('input/active_mergetab_nochim.rds')
dim(asvtab)
asvtab = t(asvtab)
dim(asvtab)
```

There are 100,259 ASVs in 13,600 samples sequenced.

```{r}
seq_ids = colnames(asvtab)
head(seq_ids)
negs = seq_ids[!startsWith(seq_ids, 'IMG')]
length(negs)
negs
```

There are 73 negative controls. I will partition them off for now.

```{r}
neg_ctl = asvtab[,negs]
seq_rem = seq_ids[!(seq_ids %in% negs)]
asvtab_no_neg = asvtab[,seq_rem]
dim(asvtab_no_neg)
```

This leaves 13,531 samples. Now let's look for any IMG ids that are missing from
the sample data sheet

```{r}
seq_only = seq_rem[!(seq_rem %in% samp_inf$Study.ID)]
length(seq_only)
sort(seq_only)
```

As of April 24, 2026, there are 49 samples that start with IMG but are missing
from the sample info sheet. Most of these end in A or B, which means they are
re-sequenced samples, but a couple don't. Let's look at those:

```{r}
asv_missing = seq_only[!grepl('[AB]$',seq_only)]
length(asv_missing)
asv_missing

(samp_inf %>% filter(grepl('D', Study.ID) |
                         grepl('discard', Notes, ignore.case = TRUE) |
                         grepl('ineligible', Notes, ignore.case = TRUE) |
                         grepl('dispose', Notes, ignore.case = TRUE))
                    %>% select(Study.ID))
samp_inf_rm
```

There are 6 sequenced samples missing from the sample info sheet. All of these
are marked DNP or otherwise removed.

Let's look at just the 43 that are duplicates, and make sure they all have
corresponding originals:

```{r}
dups = seq_only[grepl('[AB]$', seq_only)]
length(dups)
dups

dup_origs = substr(dups, 1, nchar(dups) -1)
dup_origs
n_distinct(dups)
n_distinct(dup_origs)

all(dup_origs %in% seq_rem)
```

Nope! Some are missing!

```{r}
sum(dup_origs %in% seq_rem)
```

Only 13 are in there.

```{r}
dup_origs[dup_origs %in% colnames(asvtab)]
dup_origs[!(dup_origs %in% colnames(asvtab))]
```

This is confusing. Let's see what's up with these guys in samp_info:

```{r}
sum(dup_origs %in% samp_inf$Study.ID)
```

All of them are in sample info, which is promising. Is it an A vs B thing?

```{r}
all(dup_origs[endsWith(dups, 'A')] == 
        dup_origs[!(dup_origs %in% colnames(asvtab))])
```

Okay, so the ones that end in A are lacking originals in the ASV table, and the
ones that end in B are actual duplicates that need to be merged. Cool.

```{r}
reseq = dups[endsWith(dups, 'A')]
reseq_orig = dup_origs[endsWith(dups, 'A')]

dup_origs = dup_origs[endsWith(dups, 'B')]
dups = dups[endsWith(dups, 'B')]
```

Remove the actual duplicates to their own table for now

```{r}
duptab = asvtab_no_neg[,c(dups, dup_origs)]
dim(duptab)
seq_rem = seq_rem[!(seq_rem %in% c(dups, dup_origs))]
length(seq_rem)
asvtab_no_dup = asvtab_no_neg[, seq_rem]
```

Remove the A from the column names of the non-duplicated samples so they match
the metadata

```{r}
colnames(asvtab_no_dup) = str_replace(colnames(asvtab_no_dup), 'A','')
seq_rem = colnames(asvtab_no_dup)
```

Merge the sequences of the duplicated samples. Make sure the IDs actually match.

```{r}
data.frame(dups, dup_origs)
```

They match, so we'll iterate through it:

```{r}
dedupped = matrix(nrow = nrow(duptab), ncol = length(dups))
for (i in 1:length(dups)){
    d = dups[i]
    o = dup_origs[i]
    dedupped[,i] = rowSums(duptab[,c(d,o)])
}
head(dedupped)
colnames(dedupped) = dup_origs
rownames(dedupped) = rownames(duptab)
```

Bring all the samples back together:

```{r}
asvtab_fixed = cbind(neg_ctl, asvtab_no_dup, dedupped)
dim(asvtab_fixed)
```

Let's double check that adding up the samples worked:

```{r}
asv_tots = rowSums(asvtab)
asv_tot_dedup = rowSums(asvtab_fixed)
all(asv_tots == asv_tot_dedup)

dim(asvtab)
dim(asvtab_fixed)
length(dups)
```

The rowsums are the same and there are 13 samples fewer in the fixed table than
the original table, which is the right number of duplicates. 

How many of the samples are missing from the sample info sheet?

```{r}
samp_ids = colnames(asvtab_fixed)
# head(samp_ids)
# tail(samp_ids)
# head(samp_inf)
missing = samp_ids[!(samp_ids %in% c(samp_inf$Study.ID))]
missing = missing[!(missing %in% negs)]
missing
asv_missing
```

Just the 6 that should be discarded. I'll discard them now.

```{r}
dim(asvtab_fixed)
asvtab_fixed = asvtab_fixed[,colnames(asvtab_fixed)[!(colnames(asvtab_fixed) %in% 
                                                          missing)]]
dim(asvtab_fixed)
```

Removed 6 samples.

Removing a sample may bring some ASVs down to 0 everywhere. Let's check and
remove them if necessary:

```{r}
dim(asvtab_fixed)
sum(rowSums(asvtab_fixed) == 0)
asvtab_fixed = asvtab_fixed[rowSums(asvtab_fixed) > 0,]
dim(asvtab_fixed)

dim(samp_inf)
ncol(asvtab_fixed) - nrow(samp_inf)
```

There were 25 ASVs with 0s everywhere. They have been removed, leaving us with 
100,277 ASVs in 13,585 samples.

There are 62 more columns in asvtab_fixed than in samp_inf

```{r}
all(samp_inf$Study.ID %in% colnames(asvtab_fixed))
tst = (samp_inf %>% 
           filter(!(Study.ID %in% colnames(asvtab_fixed)),
                  !grepl('D', Study.ID)))
tst

(meta_long %>% filter(Sample.ID %in% tst$Sample.ID))

all(colnames(asvtab_fixed) %in% samp_inf$Study.ID)
colnames(asvtab_fixed)[!colnames(asvtab_fixed) %in% samp_inf$Study.ID]
```

All the samples in the ASV table, but absent from the sample info sheet, are
negative controls. There are no samples in the sample info sheet that are
missing from the ASV table for no reason. All of them are DNPs.

# Write Fixed Tables

Create a maptab that is merged from the IMAGINE and Surette metadata sheets and
an asvtab where all the duplicates are merged, the sample IDs match the maptab,
and the discards are removed.

```{r}
maptab = full_join(samp_inf, meta_long, by = 'Sample.ID')
dim(maptab)
```

The negative controls aren't in the maptab, which means they will be removed
from everything downstream of this once we start using phyloseq. I'll put them
in.

```{r}
head(maptab)
plate = unlist(strsplit(negs, 'n'))[(1:length(negs)) * 2 -1]
plate
neg_map = data.frame(Sample.ID = NA,
                     Study.ID = negs,
                     Date.Received = NA,
                     Illumina.Plate.Number = plate,
                     Notes = 'negative control',
                     Site.Number = NA,
                     Participant.ID = NA,
                     Disease.Type = NA,
                     IBD.Classification = NA,
                     Timepoint = NA)
maptab = rbind(maptab, neg_map)
dim(maptab)
dim(asvtab_fixed)
```

Including negative controls, that leaves us with 13,596 samples in the maptab.
There are 13,585 in the ASV table.

```{r}
all(maptab$Study.ID %in% colnames(asvtab_fixed))
map_only = (maptab
            %>% filter(!(Study.ID %in% colnames(asvtab_fixed))))
map_only

maptab = filter(maptab, !grepl('D', Study.ID))
dim(maptab)
map_only = (maptab
            %>% filter(!(Study.ID %in% colnames(asvtab_fixed))))
map_only
```

Some of these are DNPs. Removing them gives us 13,585 samples in the maptab and
13,585 in the asv table, and they are the same 13,585! We did it!

```{r}
if (!dir.exists('data/')){
  dir.create('data', recursive = TRUE)
}

write_rds(maptab, file = 'data/merged_maptab.rds')
write_rds(asvtab_fixed, file = 'data/merged_seqtab.rds')
```

Start tracking sample/taxon counts

```{r}
dim(maptab)
dim(asvtab_fixed)

stat_df = data.frame(Step = '00_manual_qc_steps.Rmd',
                     Samples = c(nrow(maptab), ncol(asvtab_fixed)),
                     Taxa = c(NA, nrow(asvtab_fixed)),
                     File = c('data/merged_maptab.rds', 
                              'data/merged_seqtab.rds'))
stat_df
if (!dir.exists('stats')){
    dir.create('stats')
}

write.csv(stat_df, file = 'stats/track_counts.csv', row.names = FALSE)
```

